Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2024-01-08 manu-draft.qmd
intro text changes

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support growth and coverage expansion. A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in external nutrient loads. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Complementary management actions, in addition to nutrient reductions, are needed in these situations. Management actions with co=benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.

>>>>

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other critical coastal environments [@duarte2008a], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support growth and coverage expansion. A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in external nutrient loads. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Complementary management actions, in addition to nutrient reductions, are needed in these situations. Management actions with co-benefits are much more effective in estuaries that are stressed but not past critical thresholds that produce an alternative ecosystem state.
